Large-scale harmonic patterns clustering

### 2010-03-10 Paper/Paper.tex
improved
details

Working with large amount of audio data is a goal and a challenge at the
same time. One of the most common task is to find the nearest neighbor,
based on some distance, to a given segment of music. This can be viewed
as \textit{retrieval} as you are looking for some specific target
based on some query. $k$-nearest neighbors ($k$nn) algorithms have been
designed for that purpose, and recent improvements let them scale to
considerable data set sizes \cite{Datar2004}. The \textit{Shingles}
were introduced in that framework \cite{Casey2006}. A \textit{Shingle}
is a pattern of stacked chroma features. Similarity between
\textit{Shingles} coresponds to comparing harmonic patterns. For this
reason, they were used for cover or remix recognition \cite{Casey2007}.

@@

@@

\iffalse % ignored section
The original goal of the \textit{Shingles} was retrieval, i.e. find very
similar song segments based on their harmonic pattern. Part of the
challenge was to handle tons of overlapping song segments represented
by a rather long feature vector (a serie of chroma features).
The LSH algorithm proved a great solution for performing a $k$-nearest
neighbors ($k$nn) search in such a space. However, the paradigm of $k$nn has
an intrinsec scaling limit.

Finding a closest match is a tangible goal within a personal collection
of a few thousands songs. It can be used to discover cover songs or
find the seamless transitions in a playlist. However, the closest match
in millions and millions of music excerpts do not help us organize such
an amount of media. Some tasks are still important, for instance cover
recognition, and improvements in $k$nn methods are always welcome.
But it does not the describe the music space, nor does it necessarily
make good recommendations. We do not want to look for one neighbor, but
rather a cloud of potential close patterns. A similar argument in the 
recommendation domain is that precision becomes more important than recall 
passed a certain scale of data for many practical uses.

We propose to move the \textit{Shingle} idea into the larger clustering 
paradigm.
\fi % end of ignored section

The goal is to find typical patterns that can be seen as cluster centroids.
Those typical patterns help us understand a music collection from an
harmonic point of view. Segments of songs can be described in regard to their
proximity to the different centroids, thus creating some notion of distance
among the song segments. These cluster of similar patterns can have many usages:
\begin{enumerate}
\item describing the music space
\item finding potential similar songs and/or covers
\item eliminating quickly very wrong matches in a query system
\end{enumerate}

Our contribution is... large-scale clustering method,

We present a practical way to do such large-scale clustering using vector
quantization. We present clustering results of an online algorithm on a data 
set of $43$K songs, but the method applies to any data size. We show how to
select different parameters such as the size of the \textit{Shingles} 
using an encoding scheme on a test set of $4$K songs. Finally, we illustrate
the method with some specific examples.

\subsection{Definitions}
A pattern, or a patch, is a \textit{Shingle} taken from an actual song.
A code is similar to a pattern, but is the result of the clustering
algorithm and most likely does not come from an actual song.
Distortion is our measure of distance between a pattern and a code.

@@

\subsection{Echo Nest features}
We use the Echo Nest analyze API \cite{EchoNest} throughout this work.
The API gives us a chroma vector (length $12$) for every music event, 
or ``segment'', for any song we upload to their platform. 
We also get an estimate of where the beats and the bars are. 
We transform the chroma vectors per segment into per beat vector using a 
simple average. We can then stack beat vectors for a number of bars (usually 1 or 2). 
We get a fixed-size patch by resampling the number of chroma vectors. Typically,
there are $4$ beats per bar. Thus an appropriate patch size for one bar
is $12$x$4$ or $12$x$8$.

Note that we do not claim that any of these informations (segments, beats, bars)
are perfectly accurate (e.g. in \cite{Barrington2009a} authors improve on the
song segmentation features). Practise showed us that they are reasonable, 
and the size of the data set should make up for the imperfections or noise.
We also believe that patches sizes based on a number of beats or bars are more
meaningful than an arbitrary time length. More on this in the experiments
(Section \ref{sec:experiments}).

@@

\subsection{Cowbell Dataset}
We have $43,300$ songs, giving  $3,720,091$ non zero bars, a bar usually 
contains $4$ beats.

@@

@@

\section{Experiments}\label{sec:experiments}

@@

@@

% RESULTS
% 200codes, 2 bars, pSize=16, maxener, True, False, not scaled = 3.788
%                                             averaged on size = 
% 200codes, 2 bars, pSize=8, maxener, True, False, not scaled  = 3.135
%                                             averaged on size = 
% 200codes, 1 bars, pSize=4, maxener, True, False, not scaled  = 
%                                             averaged on size = 
% 200codes, 7 beats, pSize=7, maxener, True, False, not scaled = 2.771
%                                             averaged on size = 
% 200codes, 5 beats, pSize=5, maxener, True, False, not scaled = 2.397
%                                             averaged on size = 
% 200codes, 3 beats, pSize=3, maxener, True, False, not scaled = 
%                                             averaged on size = 
\begin{table}
\begin{center}
\begin{tabular}{|l|c|}
\hline
size & avg. dist \\ \hline \hline
1 bar (4) & \\
2 bars (8) & $$\\
2 bars (12) & $$\\
2 bars (16) & $$ \\ \hline
3 beats & \\
5 beats & $$\\
7 beats & $$\\ \hline
\end{tabular}
\end{center}
\caption{{Encoding score for different pattern sizes. We can select the
size of the patterns based on bars. In this case, we put the actual size
of the pattern in parentheses. If the pattern size is based on beats, the
size is the number of beats. Results are on the uspop dataset, not seen
during training.}}
\label{tab:psize}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{|l|l|c|}
\hline
\# codes & \# samples & avg. dist \\ \hline \hline
1 & $500$ & $0.195939$ \\
10 & $5$K & $0.161812$ \\
50 & $25$K & $0.147472$ \\
100 & $50$K & $0.142194$ \\
200 & $100$K & $0.137548$ \\
500 & $250$K & $0.131211$ \\
1000 & $500$K & \\ \hline
\end{tabular}
\end{center}
\caption{{Code book size.}}
\label{tab:cbsize}
\end{table}
% RESULTS:      avg_distsOLD    avg_dist
% code 1        0.195939        0.066079
% code 10       0.161812        0.045690
% code 50       0.147472        0.038456
% code 100      0.142194        0.036725
% code 200      0.137548        0.
% code 500      0.131211        0.

@@

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|}
\hline
\# codes & \# bars & avg. dist \\ \hline \hline
2 & $1$ & $0.058435$ \\
4 & $2$ & $0.054792$ \\
16 & $4$ & $0.055489$ \\
256 & $8$ & $0.054860$ \\ \hline
2 & $2$ beats* & $0.066416$ \\
4 & $1$ & $0.050712$ \\
16 & $2$ & $0.050172$ \\
256 & $4$ & $0.04989$ \\ \hline
\end{tabular}
\end{center}
\caption{{Code book size in function of the number of bars to encode.
All experiments done wtih $256 \times 500 = 128$K samples randomly chosen.
* For the $2$ beats experiment, we take the first half of every beats.
It is different from ignoring bars and taking every $2$ beats. It is more
coherent with the other settings ($1-2-4-8$ bars).}}
\label{tab:size_pattern}
\end{table}
% RESULTS:           avg_distsOLD   avg_dist
%2 codes 1 bar       0.180750       0.058435
%4 codes 2 bars      0.183631       0.054792
%16 codes 4 bars     0.184078       0.055489
%256 codes 8 bars    0.183464       0.054860
%---------------
%2 codes 2 beats     0.166840       0.046955
%2 codes 2 beats                    0.066416
% based on bars
%4 codes 1 bar       0.171495       0.050712
%16 codes 2 bars     0.173767       0.050172
%256 codes 4 bars    0.174169       0.049894

@@

\small
%\section{Acknowledgements}
%Thierry is NSERC graduate fellow, or some title like that.

>>>>

Working with large amount of audio data is a goal and a challenge at the
same time. One of the most common task is to find the nearest neighbor,
based on some distance, to a given segment of music. This can be viewed
as \textit{retrieval} as you are looking for some specific target
based on some query. $k$-nearest neighbors ($k$nn) algorithms have been
designed for that purpose, and recent improvements let them scale to
considerable data set sizes \cite{Datar2004}. The \textit{Shingles}
were introduced in that framework \cite{Casey2006}. A \textit{Shingle}
is a pattern of stacked chroma features. Similarity between
\textit{Shingles} coresponds to similarity between harmonic patterns. For this
reason, they were used for cover or remix recognition \cite{Casey2007}.

@@

We investigate a vector quantization \cite{Datar2004} (VQ) method to
cluster the \textit{Shingles}. We create a codebook of typical patterns
that have many usages:
\begin{enumerate}
\item describing the music space
\item finding potential similar songs and/or covers
\item eliminating quickly very wrong matches in a query system
\end{enumerate}

@@

In this paper, we present a practical way to do such large-scale 
clustering using online vector quantization. We present clustering results 
of an online algorithm on a data set of $43$K songs, but the method applies 
to any data size. We show how to select different parameters such as the size 
of the \textit{Shingles} using an encoding scheme on a test set of $8$K songs. 
Finally, we illustrate the method with some specific examples.

\subsection{Definitions}
A pattern, or a patch, is a \textit{Shingle} taken from an actual song.
A code is similar to a pattern, but is the result of the clustering
algorithm and most likely does not come from an actual song.
Distortion is our measure of distance between a pattern and a code
(or between two patterns, see equation \ref{eq:dist} in Section 
\ref{sec:experiments}).

@@

\subsection{Echo Nest features}
We use the Echo Nest analyze API \cite{EchoNest} throughout this work.
The API gives us a chroma vector (length $12$) for every music event, 
or ``segment'', for any song we upload to their platform. 
We also get an estimate of where the beats and the bars are. Beats span over
multiple segments, and bars span over multiple beats. 
We transform the chroma vectors per segment into per beat vectors using a 
simple average. We can then stack beat vectors for a number of bars 
(usually 1 or 2). 
We get a fixed-size patch by resampling the number of chroma vectors. Typically,
there are $4$ beats per bar. Thus an appropriate pattern size for one bar
is $12$x$4$ or $12$x$8$.

Note that we do not claim that any of these informations (segments, beats, bars)
are perfectly accurate (e.g. in \cite{Barrington2009a} authors obtain a better
song segmentation than Echonest). Practise showed us that they are reasonable, 
and the size of the data set should make up for the imperfections or noise.
We also believe that patches sizes based on a number of beats or bars are more
meaningful than an arbitrary time length. More on this in the experiments
(Section \ref{sec:experiments}).

@@

\subsection{Cowbell Dataset}
We have $43,300$ songs, giving  $3,720,091$ non zero bars, a bar usually 
contains $4$ beats. WHERE DOES THIS DATA COMES FROM? GIFT FROM EN?

@@

We try to normalize the patterns regarding the key. We roll the matrix
so the row with the more energy is always the lowest one. This can easily be
seen in Figure \ref{fig:code}. We considered the same technique for rolling
along the other axis and be invariant regarding the downbeat. However, it
seems less robust to noise. Note that we roll each pattern independentely.
Also, we imply that the pitch envelope is important rather than the actual 
pitches.

@@

\section{Experiments}\label{sec:experiments}
In this section we present general experiments that describe our method.
We start by giving some details on how we selected our training set and
how to measure error. Secondly, we experiment with different pattern sizes.
Thirdly, we look at the effect of the size of the
data set on the training. Then, we show that the patterns are not random,
which is expected from music data. Finally, we show examples of patterns,
code books and clusters.

@@

As a distance measure (or distortion measure) between patterns, we use
the average squared euclidean distance. In details, if a pattern $p1$
is composed of elements $p1_{i,j}$, and similarly for a pattern $p2$,
the distance between $p1$ and $p2$ is:
\begin{eqnarray}
  dist(p1,p2) = \sum_{i,j} (p1_{i,j} - p2_{i,j})^2 / size(p1)  \label{eq:dist}
\end{eqnarray}
We assume $p1$ and $p2$ have the same size.

@@

% RESULTS
% 200codes, 2 bars, pSize=16, maxener, True, False, not scaled = 
% 200codes, 2 bars, pSize=8, maxener, True, False, not scaled  = 0.
% 200codes, 1 bars, pSize=4, maxener, True, False, not scaled  = 
% 200codes, 7 beats, pSize=7, maxener, True, False, not scaled = 
% 200codes, 5 beats, pSize=5, maxener, True, False, not scaled = 
% 200codes, 3 beats, pSize=3, maxener, True, False, not scaled = 
\begin{table}
\begin{center}
\begin{tabular}{|l|c|}
\hline
size & avg. dist \\ \hline \hline
1 bar (4) & \\
2 bars (8) & $$\\
2 bars (12) & $$\\
2 bars (16) & $$ \\ \hline
3 beats & \\
5 beats & $$\\
7 beats & $$\\ \hline
\end{tabular}
\end{center}
\caption{{Encoding score for different pattern sizes. We can select the
size of the patterns based on bars. In this case, we put the actual size
of the pattern in parentheses. If the pattern size is based on beats, the
size is the number of beats. Results are on the uspop dataset, not seen
during training.}}
\label{tab:psize}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{|l|l|c|}
\hline
\# codes & \# samples & avg. dist \\ \hline \hline
1 & $500$ & $0.066079$ \\
10 & $5$K & $0.045690$ \\
50 & $25$K & $0.038456$ \\
100 & $50$K & $0.036725$ \\
200 & $100$K & $0.033923$ \\
500 & $250$K & $$ \\
1000 & $500$K & \\ \hline
\end{tabular}
\end{center}
\caption{{Code book size.}}
\label{tab:cbsize}
\end{table}
% RESULTS:      avg_distsOLD    avg_dist
% code 1        0.195939        0.066079
% code 10       0.161812        0.045690
% code 50       0.147472        0.038456
% code 100      0.142194        0.036725
% code 200      0.137548        0.033923
% code 500      0.131211        0.

@@

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|}
\hline
\# codes & \# bars & avg. dist \\ \hline \hline
2 & $1$ & $0.058435$ \\
4 & $2$ & $0.054792$ \\
16 & $4$ & $0.055489$ \\
256 & $8$ & $0.054860$ \\ \hline
2 & $2$ beats* & $0.066416$ \\
4 & $1$ & $0.050712$ \\
16 & $2$ & $0.050172$ \\
256 & $4$ & $0.049894$ \\ \hline
\end{tabular}
\end{center}
\caption{{Code book size in function of the number of bars to encode.
All experiments done wtih $256 \times 500 = 128$K samples randomly chosen.
* For the $2$ beats experiment, we take the first half of every beats.
It is different from ignoring bars and taking every $2$ beats. It is more
coherent with the other settings ($1-2-4-8$ bars).}}
\label{tab:size_pattern}
\end{table}
% RESULTS:           avg_distsOLD   avg_dist
%2 codes 1 bar       0.180750       0.058435
%4 codes 2 bars      0.183631       0.054792
%16 codes 4 bars     0.184078       0.055489
%256 codes 8 bars    0.183464       0.054860
%---------------
%2 codes 2 beats     0.166840       0.046955
%2 codes 2 beats                    0.066416
% based on bars
%4 codes 1 bar       0.171495       0.050712
%16 codes 2 bars     0.173767       0.050172
%256 codes 4 bars    0.174169       0.049894

% see generate_expo_codes_curves.py to generate the following figure:
\begin{figure}[htb]
\begin{center}
\includegraphics[width=.99\columnwidth]{codesize_patternsize}
\end{center}
\caption{{Possible figure to replace Table \ref{tab:size_pattern}.}}
\label{fig:size_pattern}
\end{figure}

@@

\small
%\section{Acknowledgements}
%Thierry is NSERC graduate fellow, or some title like that.
% Graham for the NMF stuff, unless is an author.
